Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-01-31 analysis/paper/paper.Rmd
methods edits

To estimate the total consumption potential (i.e., the number or amount of juvenile Chinook Salmon consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on the following inputs: predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. The predator start weight was calculated using a weight-length formula [@Parker1995] and the average length of Northern Pikeminnow caught in Deadwater Slough during our study. The average length of Northern Pikeminnow caught during the fall and spring efforts were `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively, which calculated to average starting weights of 504.4 and 670.5 g. We assumed no growth in individual Northern Pikeminnow, and so predator start and end weights are equal (i.e., we assumed the population biomass was stable). Mean daily water temperatures were summarized from 15-minute interval temperature readings available March 3, 2013 to June 14, 2021 from a gage station approximately 22 river kilometers downstream of Deadwater Slough near Shoup, Idaho [@USGS2016].

The largest uncertainty among model parameters was the proportion of the Northern Pikeminnow diet consisting of fish versus non-fish prey items (e.g., invertebrates). Therefore, we conducted nine model runs with varying proportions of fish in the diet ranging from 10-90%, in 10% increments. Energy densities for invertebrates was fixed at 3,000 J/g. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey consumed during model runs, we assume all fish prey have the same energy densities as juvenile Chinook Salmon i.e., juvenile Chinook Salmon were used as surrogates for all fish prey items. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (9 diet scenarios and 2 seasons). Each model run provided an estimate of the total amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods. To estimate the total biomass of fish potentially consumed by all Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated predator population size during each period (fall and spring).

@@

Multiplying the grams of fish consumed by an individual Northern Pikeminnow by the Northern Pikeminnow population size provides an estimate of total fish consumption potential for the population. However, the bioenergetics model runs do not differentiate juvenile Chinook Salmon from other fish prey items. Therefore, to understand potential impacts to local Chinook Salmon populations, we need to parse out the portion of total fish consumed that is juvenile Chinook Salmon. Unfortunately, estimates of the proportion of the Northern Pikeminnow fish prey that are juvenile Chinook Salmon are sparse, as well as information on the prevalence of juvenile Chinook Salmon relative to other fish prey items. Therefore, we instead consider hypothetical scenarios where Chinook Salmon are 1%, 10%, 30%, and 50% of fish prey consumed, by weight, which provide estimates of the total biomass (grams) of juvenile Chinook Salmon consumed by Northern Pikeminnow during the DSR and NRR emigration periods.

The estimate of total juvenile Chinook Salmon biomass consumed was converted to the total number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were taken from fish caught at seven rotary screw traps located throughout the Upper Salmon MPG upstream of Deadwater Slough during the fall and spring time periods.

<!-- Sawtooth Hatchery, East Fork Salmon, Pahsimeroi, Upper Lemhi, Hayden Creek, Lower Lemhi, North Fork Salmon -->

<!--
## Juvenile Chinook Salmon Abundance

Juvenile Chinook Salmon abundances were estimated at 5 rotary screw traps in the upper Salmon River basin: Sawtooth, Pahsimeroi, Hayden Creek, Upper Lemhi, and Lower Lemhi. Abundances were calculated by Idaho Department of Fish and Game at each trap which made up the production estimate for juvenile Chinook Salmon in the Upper Salmon Basin (Table X) (McClure et al. 2020).
-->

@@

Using hook-and-line angling, we caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 efforts, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` fish. The CPUE, measured as number of Northern Pikeminnow caught per angler hour, was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021 (Figure \@ref(fig:cpue-fig)). During the mark-recapture efforts, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019 which included `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals. There were an additional `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020 consisting of `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` (Table \@ref(tab:abund-tab)). In both cases, the multiple census estimators were larger than the single census estimators. Confidence intervals for the fall 2020 effort were more varied, likely due to fewer recaptures during that year.

>>>>

To estimate the total consumption potential of Northern Pikeminnow in Deadwater Slough during the peaks of fall DSR and spring NRR emigrations, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on the following inputs: predator and prey energy densities, predator start and end weights, and water temperatures. Separate models were run for the fall period (September 15 - November 30) and the spring period (March 1 - May 31) to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG.

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because were were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average length of Northern Pikeminnow caught in Deadwater Slough during our study was converted to weight using a weight-length formula from @Parker1995. This resulted in average spring and fall starting eights of 504.4 and 670.5 g for Northern Pikeminnow with average lengths of `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 RKM downstream of Deadwater Slough.

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted separate model runs that assume different proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey. 14 models were run (7 diet scenarios and 2 seasons), each providing an estimate of the total grams of fish consumed by an individual Northern Pikeminnow (***table reference***). To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population size during fall and spring.

@@

The proportion of Chinook Salmon prey relative to other fish prey in the Northern Pikeminnow diet is also unknown. However, there is some evidence that juvenile salmonids are by far the most consumed fish prey [@Zimmerman1999; @Shively1996]. Moreover, it is estimated that Chinook Salmon make up 64.2%, 29.3%, and 49.3% of the fish prey consumed by Northern Pikeminnow in the Columbia River below Bonneville Dam, Columbia River reservoirs, and in the lower Snake River, respectively [@Zimmerman1999]. Because the bioenergetics model does not differentiate juvenile Chinook Salmn from other fish prey items, we performed a sensitivity analysis to assess potential impacts of Northern Pikeminnow predation on local Chinook Salmon populations. Using values similar to @Zimmerman1999 we modeled three diet scenarios where Chinook Salmon comprised 30%, 50%, and 65% of the fish prey, by weight, in the Northern Pikeminnow diet during the fall DSR and spring NRR emigration periods. In this analysis, we assume 60% of the Northern Pikeminnow's total diet is comprised of fish, representing the median in our range of modeled diets.

The estimate of total juvenile Chinook Salmon biomass consumed was then converted to the total estimated number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods.

Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is often adult returns. We assessed the impacts on adult returns by estimating the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. This calculation was made by multiplying the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

@@

Using hook-and-line angling, we caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 surveys, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` fish. The CPUE was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021 (Figure \@ref(fig:cpue-fig)). During the mark-recapture events, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019, including `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals and `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020, including `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` (Table \@ref(tab:abund-tab)).For both estimates, the multiple census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (***figure reference***).
